---
title: "Peatlands and Torbiere"
---

### Why Peatlands Matter

Peatlands are among the most quietly remarkable landscapes on Earth. Formed over thousands of years from partially decomposed plant material accumulating in waterlogged conditions, they lock away vast amounts of carbon, filter water, and support specialised plants and animals found almost nowhere else. But their significance has never been purely ecological. Across very different cultures and histories, peatlands have long been places of livelihood, myth, and memory — sites of turf-cutting and grazing, but also of folklore, fear, and reverence for what lies beneath the surface. Ireland and Italy tell very different stories about their peatlands, shaped by entirely different geographies, land-use histories, and relationships to language — yet in both places, these landscapes are now being rethought: not only as carbon stores to be protected, but as living archives of ecological and cultural knowledge.


## Corveen Bog, County Donegal, Ireland

Corveen Bog is an upland blanket bog on the north-facing slopes of **Tievalehid Mountain**, in the **Gweedore district of northwest Donegal**, at altitudes between 120m and 430m and exposed to the full force of the Atlantic. It is designated a **Natural Heritage Area** and is recognised as the only reasonably intact stretch of blanket bog remaining in the district — a landscape type that, while once widespread across Ireland and Britain, has been dramatically reduced by overgrazing and afforestation.

```{r}
#| echo: false

library(leaflet)

leaflet(
  options = leafletOptions(
    zoomControl = FALSE
  )
) |>

  addTiles(
    attribution = "&copy; OpenStreetMap contributors"
  ) |>

  # Approximate footprint of Corveen Bog
  addPolygons(
    lng = c(
      -8.25, -8.23, -8.20, -8.17, -8.15,
      -8.14, -8.16, -8.19, -8.22, -8.25
    ),
    lat = c(
      55.09, 55.11, 55.12, 55.11, 55.09,
      55.07, 55.06, 55.07, 55.08, 55.09
    ),
    color = "#163B32",
    weight = 2,
    fillColor = "#DCD0FF",
    fillOpacity = 0.35
  ) |>

  # Corveen Bog label
  addLabelOnlyMarkers(
    lng = -8.205,
    lat = 55.091,
    label = "Corveen Bog",
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "center",
      textOnly = TRUE,
      className = "study-label"
    )
  ) |>

  # Gortahork
  addLabelOnlyMarkers(
    lng = -8.225,
    lat = 55.065,
    label = "Gortahork",
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "bottom",
      textOnly = TRUE,
      className = "city-label"
    )
  ) |>

  # Bunbeg
  addLabelOnlyMarkers(
    lng = -8.300,
    lat = 55.070,
    label = "Bunbeg",
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "left",
      textOnly = TRUE,
      className = "city-label"
    )
  ) |>

  # Donegal
  addLabelOnlyMarkers(
    lng = -8.110,
    lat = 54.654,
    label = "Donegal",
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "right",
      textOnly = TRUE,
      className = "city-label"
    )
  ) |>

  fitBounds(
    lng1 = -8.45,
    lat1 = 54.95,
    lng2 = -7.90,
    lat2 = 55.20
  ) |>

  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      imperial = FALSE
    )
  )
```


The bog supports a striking diversity of plant life: black bog-rush, deergrass, hare's-tail cottongrass, and purple moor-grass across its drier plateau, with well-developed pool systems colonised by sphagnum mosses, bladderwort, and bogbean. Rarer species include the greater sundew, bog asphodel, and the long-stalked yellow-sedge — of particular note here as it is otherwise scarce in County Donegal. Corveen also provides habitat for several species on Ireland's Red Data List, including merlin, golden plover, red grouse, and Irish hare.

Historically shaped by sheep grazing and turf-cutting — both mechanised and by hand — the site still bears the marks of drainage channels dug for turbary plots. Today, like many Irish peatlands, Corveen sits at the intersection of conservation and cultural continuity: close to **Gaoth Dobhair**, one of Ireland's strongest Irish-speaking communities, its restoration is entangled with the everyday survival of the Irish language itself.

## Parco dell'Alto Appennino Modenese, Modena, Italy

High in the **Tuscan–Emilian Apennines**, in the cluster of glacial lakes and peatlands around **Lago Santo** and **Lago Baccio**, this project focuses on a small group of high-altitude peat bogs within the Natura 2000 site "Monte Rondinaio, Monte Giovo", part of the Alto Appennino Modenese Regional Park. Lago Santo itself, at 1,501 metres, is one of the largest natural lakes in the northern Apennines, its steep glacial basin ringed by rock and scree from Monte Giovo. It is a well-loved destination for hikers and anglers — but the smaller, quieter peatlands nearby tell a different, less-visited story.


```{r}
#| echo: false

library(leaflet)

leaflet(
  options = leafletOptions(
    zoomControl = FALSE
  )
) |>

  addTiles(
    attribution = "&copy; OpenStreetMap contributors"
  ) |>

  # Approximate footprint of the Alto Appennino Modenese
  addPolygons(
    lng = c(
      10.48, 10.50, 10.54, 10.58, 10.62,
      10.67, 10.72, 10.77, 10.82, 10.83,
      10.81, 10.77, 10.73, 10.68, 10.63,
      10.58, 10.53, 10.49
    ),
    lat = c(
      44.23, 44.24, 44.22, 44.20, 44.21,
      44.23, 44.23, 44.22, 44.20, 44.18,
      44.16, 44.14, 44.13, 44.12, 44.12,
      44.13, 44.15, 44.18
    ),
    color = "#163B32",
    weight = 2,
    fillColor = "#DCD0FF",
    fillOpacity = 0.35
  ) |>

  # Park label
  addLabelOnlyMarkers(
    lng = 10.66,
    lat = 44.18,
    label = "Parco dell'Alto Appennino Modenese",
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "center",
      textOnly = TRUE,
      className = "study-label"
    )
  ) |>

  fitBounds(
    lng1 = 10.35,
    lat1 = 44.05,
    lng2 = 11.55,
    lat2 = 44.82
  ) |>

  addScaleBar(
    position = "bottomleft",
    options = scaleBarOptions(
      imperial = FALSE
    )
  )
```


Just above Lago Santo, on the path to the Passo della Boccaia, the **Torbiera Boccaia** occupies a gently sloping shelf within beech woodland, fed year-round by a permanent spring. Its southern reaches support alkaline fen vegetation dominated by broad-leaved cottongrass, while sedge communities dominate elsewhere, alongside marsh violet and butterwort. Further along the high valley of the Torrente Tagliole, **Torbiera Le Gore** — one of the most extensive peatlands in the northern Apennines — sits below Lago Turchino, its glacially formed basin threaded with streams that shape distinct plant communities across the site, including rush and sedge fens and, along its watercourses, tufa-forming spring vegetation. A little further downstream, **Torbiera Le Lamacce** forms across two linked terraces fed by springs and by the outflow of Le Gore, hosting similarly rich sedge and cottongrass communities alongside rarer plants tied to calcareous spring-fed ground. Common frogs breed across all three sites.

Unlike Corveen, these Apennine peatlands have been largely untouched by industrial extraction, but — much like Emilian, the endangered language spoken in the communities around them — they remain peripheral in public and academic attention. Recent restoration work in the wider Alto Appennino Modenese area, aimed at maintaining and expanding these wetlands and the carbon they hold, reflects a growing regional recognition of their ecological value, even as their cultural and linguistic significance remains, for now, largely untold.
